---
title: "Homework 1: Gapminder"
output: github_document
---

&nbsp;
In this assignment I will look at the  [**gapminder**](https://cran.r-project.org/web/packages/gapminder/index.html) data set, let us begin by seeing what variables are available,

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
library(gapminder)
library(ggplot2)
names(gapminder)
```

and their respective formats.

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
str(gapminder)
```

Let us now have a small look at the actual data itself,

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
head(gapminder)
```

and a few descriptive statistics.

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
summary(gapminder)
```

On seeing what the data set contains we can now ask specific questions. For instance, has *life expectancy* changed from the early *1950s to 2007*. 

<center> <h3>Life Expectancy Over Decades </h3> </center>
```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
plot(lifeExp ~ year, gapminder)
```

Looking at this scatter plot a rough trend does seem evident. Let us try and better visually summarize this by creating a figure where mean life expectancy is plotted for each year.

&nbsp;

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
library(gapminder)
library(ggplot2)

gp <- gapminder
gp$Continent <- as.factor(gp$continent)

removal = theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(), 
                panel.background = element_blank(), 
                legend.key = element_blank(),
                axis.line = element_line(color = "black"))

LifeByYear = ggplot(gp, aes(year, lifeExp)) + removal
LifeByYear + stat_summary(fun.y = mean, geom = "point", size = 4) +
    stat_summary(fun.y = mean, geom = "line", size = 1.25) +
    stat_summary(fun.data = mean_se , geom = "errorbar", width = 2) +
    xlab("Year") +
    ylab("Life Expectancy") +
    coord_cartesian(ylim = c(40,80), xlim = c(1952,2007))+
    scale_y_continuous(breaks=seq(40,80,5)) +
    scale_x_continuous(breaks=seq(1952,2007,5)) +
    theme(axis.text.x = element_text(colour="grey20",size=12,angle=0,face="plain"),
        axis.text.y = element_text(colour="grey20",size=12,angle=0,face="plain"),  
        axis.title.x = element_text(colour="grey20",size=15,angle=0,face="plain"),
        axis.title.y = element_text(colour="grey20",size=15,angle=90,face="plain"))
```

<b><i>note.</b></i>  Error bars represent +/- 1 one standard error of the mean.

As we can see life expectancy has increased each year, resulting in gain of roughly 15 years since study onset. A more nuanced distinction however may be made by considering the growth rates of different *continents*. 

&nbsp;

<center> <h3>Life Expectancy Across Continents</h3> </center>

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}

LifeByYear2 = ggplot(gp, aes(year, lifeExp, group = Continent)) + removal

LifeByYear2  +
  stat_summary(fun.data = mean_se , geom = "errorbar", width = 2) +
  stat_summary(fun.y = mean, geom = "line", aes(group = Continent, color = Continent), size = 1.25) +
  xlab("Year") +
  ylab("Life Expectancy") +
  coord_cartesian(ylim = c(40,80), xlim = c(1952,2007))+
  scale_y_continuous(breaks=seq(40,80,5)) +
  scale_x_continuous(breaks=seq(1952,2007,5)) +
  theme(axis.text.x = element_text(colour="grey20",size=12,angle=0,face="plain"),
        axis.text.y = element_text(colour="grey20",size=12,angle=0,face="plain"),  
        axis.title.x = element_text(colour="grey20",size=15,angle=0,face="plain"),
        axis.title.y = element_text(colour="grey20",size=15,angle=90,face="plain"))
```
<b><i>note.</b></i> Error bars represent +/- 1 one standard error of the mean.

&nbsp;

Visually a trend seems apparent. Continents like Asia and America had  larger increases in life expectancy than did Europe and Oceania. Interestingly Asia and America both also had a lower life expectancy at study start. Perhaps then, continents that started with a lower life expectancy had an opportunity for greater gains over the years. A ranking of average life expectancy increase is provided in the table below.

&nbsp;

|    **Continent**    | **Mean Life Expectacy in 1952** | **Mean Life Expectancy in 2007 **| **Life Expectancy Increase**| **Relative ranking**
  |----|--------------------|---------------------|----------------|---|
  |Africa          | 39.14 |54.81| 15.67|3|
  |Americas        | 53.28 |73.81| 20.53|2|
  |Asia            | 46.31 |70.73|24.42|1|
  |Europe          | 64.41 |77.65|13.24|4|
  |Oceania         | 69.26 |80.72|11.46|5|


Overall it would seem that continents with lower life expectancy saw greater gains over the years than continents with already high life expectancy. Somewhat puzzling however is that Africa although having the lowest life expectancy only showed an intermediate increase. Why might this be? One notable difference is that unlike the other continents Africa did not have nearly as pronounced an increase in *gross domestic product (GDP)* over the years. 

&nbsp;

<center> <h3>GDP Across Continents</h3> </center>

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.align="center"}
GDPByYear = ggplot(gp, aes(year, gdpPercap, group = Continent)) + removal

GDPByYear  +
  stat_summary(fun.y = mean, geom = "line", aes(group = Continent, color = Continent), size = 1.25) +
  xlab("Year") +
  ylab("GDP (per capita)") +
  coord_cartesian(ylim = c(0,35000), xlim = c(1952,2007))+
  scale_y_continuous(breaks=seq(0,35000,5000)) +
  scale_x_continuous(breaks=seq(1952,2007,5)) +
  theme(axis.text.x = element_text(colour="grey20",size=12,angle=0,face="plain"),
        axis.text.y = element_text(colour="grey20",size=12,angle=0,face="plain"),  
        axis.title.x = element_text(colour="grey20",size=15,angle=0,face="plain"),
        axis.title.y = element_text(colour="grey20",size=15,angle=90,face="plain"))
```
<b><i>note.</b></i> Error bars were not included to avoid an overly cluttered figure.

Whether Africa did not see as steady an increase in life expectancy is attributable to a halted GDP  would seem tenable.  

According to more recent 2016 statistics provided by the [**International Monetary Fund**](http://www.imf.org/external/datamapper/NGDPD@WEO/OEMDC/ADVEC/WEOWORLD/AFQ) Africa currently has a similar GDP as in the 2007 measurement. One might expect and hypothesize that as GDP has not greatly changed since the 2007 measurement, perhaps neither has life expectancy. 

&nbsp;